---
title: "Chapter 2 - Inference Using t-Distributions"
author: "The Statistical Sleuth"
date: "February 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(Sleuth3)
library(lattice)
```
## Inference Using t-Distributions

### One-Sample t-Tools and the paired t-test

If a population has a mean $\mu$ and a standard deviation $\sigma$, then the mean of the sampling distribution of the average is also $\mu$, the standard deviation of the sampling distribution is $\sigma/\sqrt{n}$, and the shape of the sampling distribution is more nearly normal than is the shape of the population distribution. 
The standard deviation in the sampling distribution of an average, denoted by $SD(\bar{Y})$, is the typical size of $(\bar{Y}-\mu)$, the error in using $\bar{Y}$ as an estimate of $\mu$. This standard deviation gets smaller as the sample size increases. 

### The Standard Error of an Average in Random Sampling

The *standard error* of any statistic is an estimate of the standard deviation in its sampling distribution. Associated with standard error is a measure of the amount of information used to estimate variability, called its *degrees of freedom*. Degrees of freedom are measured in units of "equivalent number of independent observations."

The formula for the standard deviation of the average in a sample size  $n$ is $\sigma/\sqrt{n}$, so if $s$ is the sample standard deviation, the standard error for the average is $$SE(\bar{Y}) = \frac{s}{\sqrt{n}}, d.f. = (n-1))$$

The degrees of freedom in a single sample standard deviation is always one less than the sample size. 

### The t-Ratio Based on a Sample Average

The ratio of an estimate's error to the anticipated size of its error provides a convenient basis for drawing inferences about the parameter in question. 

